--- 
title: "Quant Bio Modules"
author: "Kim Cuddington, Andrew Edwards, Brian Ingalls"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "A set of modules on quantitaitve topics for biology graduates"
---

#  Prerequisites

All chapters of this book are stored as .Rmd files on github. The whole book can be compiled to various formats using the **bookdown** package which can be installed from CRAN or Github.


```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


```{r setup, include=FALSE}
###--- Update the knitr working directory (R works from the project folder, knitr works in the rmd folder)
dd <- getwd()
knitr::opts_knit$set(root.dir= paste(dd,'/../../')) 


```
Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

