---
title: "An Introduction to Modern Tools for Quantitative Biology"
author: "Kim Cuddington, Andrew M. Edwards, Brian Ingalls"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "An Introduction to Modern Tools for Quantitative Biology"
---

#  Welcome!

Quantitative skills are essential for today's biologists. We consider such
skills to include the computational, statistical and mathematical techniques used to study life and
living organisms, including aspects of big data, transparency and reproducibility in science.
However, the breadth of quantitative techniques now employed in biology make it very likely
that there may be no suitable local expertise within a student's home
institution or an applied biologist's workplace.

A recent workshop on [Quantitative Education in Life Science Graduate
Programs](http://www.nimbios.org/workshops/WS_quantedu){target="_blank"}
identified short online training modules as a
way graduate students might acquire the relevant skills. This learner-centered approach allows
students and faculty advisors to select subtopics that are directly relevant to the students’
program, thus simultaneously providing more targeted training and reducing the time
commitment that would be involved in more generalized formal courses. This flexibility is
particularly important given the breadth of quantitative skills used in various research areas and
the differences in experience of students and advising faculty.

This e-book consists of five independent modules, aimed at biology graduate
students and at applied biologists working in academia, government
agencies and private organisations. A key theme is that while the techniques presented are
from various disciplines (such as computer science, statistics and mathematics)
we aim to introduce the key elements in a way suitable for a biological
audience.

The five modules are:

* Introduction to Git and GitHub (Andrew Edwards) -- covers the tools widely used to share code,
  collaborate with colleagues and create a transparent record of research.

* Introduction to R Markdown (Andrew Edwards) -- demonstrates how to create dynamic documents
  that are fully reproducible, with a clear link between the underlying R code
  and the resulting figures, tables and results.

* Introduction to multivariate analysis (Kim Cuddington) -- discusses multivariate quantitative
  methods that are used when there is more than one response (or measurement)
  which we are trying to understand by using more than one feature or
  characteristic of the system.

* Introduction to classification and machine learning (Kim Cuddington) -- continues explorations
  of multivariate data but using machine learning approaches. Classification is
  the task of assigning data objects, such as sites, species or images, to
  predetermined classes. Machine learning is an application of artificial
  intelligence, which allows algorithms to become more accurate at predicting
  outcomes without being explicitly programmed to do so. The algorithm itself
  completes the tasks of model specification, evaluation and improvement.

* Optimization (Brian Ingalls) -- optimization is the act of identifying the extreme (cheapest,
  tallest, fastest, ...) over a collection of possibilities. Biological
  applications include optimal manipulation of biological systems (e.g. optimal
  harvesting or optimal drug dosing), optimal design of biological systems
  (e.g. robust synthetic genetic circuits) and optimal experimental design. This
  module introduces optimization methods in R.

The modules can be used independently (i.e., you don't have read the whole
e-book!). Pick and choose want you need. Each module is designed to quickly get
you up and running on the topic of interest in 3-5 hours. As such, these
materials should be thought of a basic introductions. We provide pointers to
more advanced materials, but our aim is to "jump-start" your use of these tools.

On the other hand, the materials are not designed for absolute beginners. Our
target audience is graduate students and professionals in biology. All modules
(except the Git and GitHub one) use the programming language R, for which we
assume prior familiarity. For
material on statistical analyses we assume the standard background in univariate
statistics.

## About the authors

Kim Cuddington is an Associate Professor at the University of Waterloo in the Department
of Biology, with a cross-appointment to Applied Mathematics. She has worked on quantitative
education for biology students and has training in instructional design
(University of Waterloo, Center for Teaching Excellence). She has developed
courses on quantitative analysis in biology and previously converted two courses
to online delivery.

Andrew Edwards is a Research Scientist with the Department of Fisheries and Oceans Canada (DFO) at the
Pacific Biological Station in Nanaimo, British Columbia, and holds an Adjunct Professor position in the
Department of Biology at the University of Victoria. He has
previously developed in-person workshops on Git, GitHub and R Markdown (tools
that are widely used in DFO), and co-developed a
quantitative biology course at Dalhousie University. Current work includes
developing methods for fitting size spectra to data, conducting fisheries stock
assessments (including for Pacific Hake, the largest groundfish stock off the
west coast of North America), and recent guidelines to [using mathematical notation
in ecology](https://doi.org/10.1111/2041-210X.13105){target="_blank"}.

Brian Ingalls is an Associate Professor in the Department of Applied Mathematics at the University of Waterloo, cross-
appointed to the Departments of Biology and Chemical Engineering. He is the author of
Mathematical Modeling in System Biology: An Introduction (2013), and has previously taught
undergraduate and graduate courses on quantitative techniques in biology (including
Computational Modelling of Cellular Systems and Mathematical Cell Biology).

## Acknowledgments

We acknowledge support from the Government of Ontario through a grant from
eCampusOntario, and the support of DFO.

Several students helped with the project and provided valuable feedback. These
include TODO

The Git and GitHub module builds upon workshop materials that were originally
developed with Chris Grandin (DFO), who AME also thanks for assistance with the
module.


This is a living document and so revisions to current chapters, and the addition
of new chapters, should be anticipated. TODO - sure you want to say that??

## Open Science

TODO - need the links if we're going to say this. Tricky because of the six
separate repositories. Could just give a link to the README of the main one and
not scare people off with early talk of tex.

In the spirit of open science, all chapters of this book are stored as .Rmd files on github. The whole book can be compiled to various formats using the **bookdown** package which can be installed from CRAN or Github.


```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
## Feedback

If you use these materials please take some time to let us know how we might improve [Survey](https://forms.gle/vLkcZq7NxPb2ajTQ6).
